---
title: "Reproducible Research: Activity Monitoring Data"  
output: 
  html_document:
    keep_md: true
---

## Data

Dataset: Activity monitoring data [52K] (activity.zip)  
Source: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip  

NOTE: activity.zip dataset extracted to ./data/activity.csv

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

The variables included in this dataset are:  

* steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)  
* date: The date on which the measurement was taken in YYYY-MM-DD format  
* interval: Identifier for the 5-minute interval in which measurement was taken   

## Loading and preprocessing the data

1. Load the data (i.e. read.csv())  

2. Process/transform the data (if necessary) into a format suitable for your analysis

```{r}

## load plyr/dplyr libraries to work with data
library(dplyr)

## read data 
## NOTE: activity.zip dataset extracted to ./data/activity.csv
cols = c("numeric","character","numeric")
activity_data <- read.csv("./data/activity.csv", 
                              header = TRUE,
                              na.strings = "NA",
                              colClasses = cols)

str(activity_data)
```

## What is mean total number of steps taken per day?

NOTE: NA values ignored

1. Make a histogram of the total number of steps taken each day  

```{r figure1, fig.height=7}

## remove NAs from data 
activity_removena <-  na.omit(activity_data)

## convert to data frame table objet
activity_removena_tbl <- tbl_df(activity_removena)
str(activity_removena_tbl)

## group by date and then sum steps
by_date <- group_by(activity_removena_tbl, date)
activity_removena_date_steps <- summarize(by_date, 
                                          "total.steps" = sum(steps),
                                          "mean.steps" = mean(steps),
                                          "median.steps" = median(steps))
str(activity_removena_date_steps)

## print histogram of total number of steps per day
hist(activity_removena_date_steps$total.steps,
     col = "blue",
     xlab = "Total Number of Steps Per Day",
     main = "Histogram of Total Number of Steps Per Day (NAs Removed)",
     xlim = c(0,25000),
     ylim = c(0,30),
     axes = TRUE)
box()

```

2. Calculate and report the mean and median total number of steps taken per day

```{r}
## print summary of total numbers of steps
summary(activity_removena_date_steps$total.steps)

## report the mean and medium total number of steps per day
print(paste("Mean:",
            mean(activity_removena_date_steps$total.steps)))
print(paste("Median:",
            median(activity_removena_date_steps$total.steps)))
```

**The mean total number of steps taken per day is: `r mean(activity_removena_date_steps$total.steps)` and the median total number of steps taken per day is: `r median(activity_removena_date_steps$total.steps)`**  

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)  

```{r figure2, fig.height=7}

## group by interval and then average (mean) steps
by_interval <- group_by(activity_removena_tbl, interval)
activity_removena_interval_steps <- summarize(by_interval, 
                                          "average.steps" = mean(steps))
str(activity_removena_interval_steps)

## make time series plot (type = "l") of interval (x-axis) vs. average steps across all days (y-axis)
plot(activity_removena_interval_steps$interval, 
     activity_removena_interval_steps$average.steps,
     type = "l",
     main = "Time Series Plot of Interval vs. Average Number of Steps",
     xlab = "5-Minute Interval",
     ylab = "Average (Mean) Number of Steps Across All Days",
     col = "blue",
     xlim = c(0,2500),
     ylim = c(0,250))

```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}

## report interval with max average steps
print(paste("Interval:", 
      activity_removena_interval_steps[which.max(activity_removena_interval_steps$average.steps),1]))
print(paste("Max Steps:", 
      activity_removena_interval_steps[which.max(activity_removena_interval_steps$average.steps),2]))
```

**The `r activity_removena_interval_steps[which.max(activity_removena_interval_steps$average.steps),1]` 5-minute interval, on average across all days, contains the max number of steps (`r activity_removena_interval_steps[which.max(activity_removena_interval_steps$average.steps),2]`)**

## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}

## count and remport number of NA values
print(paste("NAs:", sum(is.na(activity_data))))

```

**Total number of missing values in dataset is: `r sum( is.na( activity_data ) )`**

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

**STRATEGY: Replaced NA values with average/mean for that 5-min interval across all days (previously calculated above)**

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}

## convert data to data frame table object
activity_tbl_replaceNA <- tbl_df(activity_data)
str(activity_tbl_replaceNA)

## loop over each row of activity data
for (i in 1:17568) {
     ## check if step value is NA
     if (is.na(activity_tbl_replaceNA[i,1])) {
        ## if NA, replace NA with average steps from interval across all days (calculated above)
        interval2 <- activity_tbl_replaceNA[i,3]
        newval <- filter(activity_removena_interval_steps, interval == as.numeric(interval2))[2]
        activity_tbl_replaceNA[i,1] <- newval
     }
}

## new data set with filled-in NA values is the "activity_tbl_replaceNA" data frame table object
str(activity_tbl_replaceNA)

```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r figure3,fig.height=7}

## group by date and then sum steps
by_date_replaceNA <- group_by(activity_tbl_replaceNA, date)
activity_tbl_replaceNA_steps <- summarize(by_date_replaceNA, 
                                          "total.steps" = sum(steps),
                                          "mean.steps" = mean(steps),
                                          "median.steps" = median(steps))
str(activity_tbl_replaceNA_steps)

## print histogram of total number of steps per day
hist(activity_tbl_replaceNA_steps$total.steps,
     col = "red",
     xlab = "Total Number of Steps Per Day",
     main = "Histogram of Total Number of Steps Per Day (NAs Filled-In)",
     xlim = c(0,25000),
     ylim = c(0,40),
     axes = TRUE)
box()

```

**These shape of the histogram with NAs filled-in is consistent when the previous histogram with NAs removed; however, the frequency values for the middle group has increased.**

```{r}
## print summary of total numbers of steps
summary(activity_tbl_replaceNA_steps$total.steps)

## report the mean and medium total number of steps per day
print(paste("Mean:",
            mean(activity_tbl_replaceNA_steps$total.steps)))
print(paste("Median:",
            median(activity_tbl_replaceNA_steps$total.steps)))
```

**The mean total number of steps taken per day is: `r mean(activity_tbl_replaceNA_steps$total.steps)` and the median total number of steps taken per day is: `r median(activity_tbl_replaceNA_steps$total.steps)`**

**The mean total number of steps taken per day did not change betwwen NA values being removed vs. filled-in. The median total number of steps slightly increased betwwen NA vlues being removed vs. filled-in.**

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}

## make copy of dataset to modify 
activity_tbl_replaceNA_weekdays <- activity_tbl_replaceNA
str(activity_tbl_replaceNA_weekdays)

## convert date column (character) to date object POSIXlt/POSIXct
activity_tbl_replaceNA_weekdays$date <- strptime(activity_tbl_replaceNA_weekdays$date,
                                                  format="%Y-%m-%d")

## add a column with the name of the day of the week
activity_tbl_replaceNA_weekdays$day <- weekdays(activity_tbl_replaceNA_weekdays$date) 

## define date names and weekday/weekend classification
date_types <- data.frame("day" = c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday"),"day.type" = c("weekday","weekday","weekday","weekday","weekday","weekend","weekend"))
date_types$day <- as.character(date_types$day)

## add "weekday" and "weekend" factor column by join-ing tables
activity_tbl_replaceNA_weekdays_weekends <- plyr::join(activity_tbl_replaceNA_weekdays, 
                                                   date_types, 
                                                   by = 'day')

## "activity_tbl_replaceNA_weekdays_weekends" is the new dataset with weekend/weekday column
## join function outputs data.frame, use tbl_df to convert back to data frame table object
activity_tbl_replaceNA_weekdays_weekends <- tbl_df(activity_tbl_replaceNA_weekdays_weekends)
str(activity_tbl_replaceNA_weekdays_weekends)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was creating using simulated data:

```{r figure4, fig.height=7}

## convert date column back to character values because dplyr function do not work well with POSIXlt
## R crashes if dplyr function used on a data frame table containing POSIXlt column
activity_tbl_replaceNA_weekdays_weekends$date <- as.character(activity_tbl_replaceNA_weekdays_weekends$date)

## group data by interval and day.type, then average (mean) steps
by_interval_daytype <- group_by(activity_tbl_replaceNA_weekdays_weekends, interval, day.type)
activity_tbl_replaceNA_weekdays_weekends_interval_average_steps <- summarize(by_interval_daytype, 
                                          "average.steps" = mean(steps))
str(activity_tbl_replaceNA_weekdays_weekends_interval_average_steps)

## make a ggplot compairing weekend vs. weekday average steps (y-axis) vs. interval (x-axis)
## base ggplot plot
library(ggplot2)
g <- ggplot(activity_tbl_replaceNA_weekdays_weekends_interval_average_steps, 
            aes(x=interval, y=average.steps))
## add layers
g2 <- g + geom_line(col = "blue") + 
          facet_grid(day.type ~ .) +
          labs(x = "5-Minute Interval") +
          labs(y = "Average (Mean) Number of Steps Across All Days") +
          labs(title = "Time Series Plot of Interval vs. Aerage Number of Steps") +
          theme_bw()
# display plot
g2
```

**Yes, the activity pattern between weekdays and weekends are different.**
