---
title: "Severe Weather Events with Greatest Fatalities, Injuries, and Economic Damage in the United States 1950-2011"
output: 
  html_document:
    keep_md: true
---

## Synopsis

Severe weather data (Source: NOAA) from 1950 to 2011 were analyzed to determine the weather events that cause the greatest population harm (fatalities and injuries) and economic damage (property and crop). The top 5 fatality causing events were TORNADO (5633), EXCESSIVE HEAT (1903), FLASH FLOOD (978), HEAT (937), and LIGHTNING (816). The top 5 injury causing events were TORNADO (91346), TSTM WIND (6957), FLOOD (6789), EXCESSIVE HEAT (6525), and LIGHTNING (5230). The top 5 property damage causing events were FLOOD ($1.447e+11), HURRICANE/TYPHOON ($6.931e+10), STORM SURGE ($4.332e+10), FLASH FLOOD ($1.682e+10), and HAIL ($1.574e+10). The top 5 crop damage causing events were DROUGHT ($1.397e+10), FLOOD ($5.662e+09), RIVER FLOOD ($5.029e+09), ICE STORM ($5.022e+09), and HAIL ($3.026e+09). Conclusions: (1) For greatest population health impact, future work/research/preparedness/aid should focus on Tornados (2) For greatest economic impact, future work/research/preparedness/aid should focus on Floods and Droughts. 

## Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Methods

Analysis was preformed on:

* R: 3.1.1 
* Rstuido: Version 0.98.1073 
* Operating System: Mac OS X 10.9.5 
* Hardware: Macbook Pro, 2.2 ghz Intel Core 2 Duo, 3 GB RAM 

NOAA Storm Data Source: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

## Data Processing 

Loaded libraries. Dplyr was used to sort/analyze dataset. 

```{r}

## load libraries --------------------------------------------------------------------------

library(stringr)
library(dplyr)

```

Loaded data and then made transformations (capitalized and removed whitespace) to clean up the inconsistencies/errors in the dataset. 

```{r, cache=TRUE}

## load data -------------------------------------------------------------------------------

data_extracted <- bzfile("repdata-data-StormData.csv.bz2")
storm_data <- read.csv(data_extracted, header = TRUE)

## data transformations --------------------------------------------------------------------

## capitalize column data to be consistant and remove whitespace to clean up data
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, toupper)
storm_data$EVTYPE <- str_trim(storm_data$EVTYPE)
storm_data$PROPDMGEXP <- sapply(storm_data$PROPDMGEXP, toupper)
storm_data$CROPDMGEXP <- sapply(storm_data$CROPDMGEXP, toupper)

## convert to table data frame
storm_data_tbl <- tbl_df(storm_data)

## print dataset
storm_data_tbl
```

Sorted data by weather event and summed the number of fatalities and injuries. Ordered the data from highest to lowest. 

```{r, cache=TRUE}

## Fatalities and Injuries ----------------------------------------------------------

## group by weather type, EVTYPE
by_EVTYPE <- group_by(storm_data_tbl, EVTYPE)

## sum fatalities and injuries by EVTYPE
tbl_FAT <- summarize(by_EVTYPE, 
                     "TOTAL.FATALITIES" = sum(FATALITIES))
tbl_INJ <- summarize(by_EVTYPE, 
                     "TOTAL.INJURIES" = sum(INJURIES))

## arrange data in decending order
tbl_FAT_order <- arrange(tbl_FAT, 
                         desc(TOTAL.FATALITIES))
tbl_INJ_order <- arrange(tbl_INJ, 
                         desc(TOTAL.INJURIES))

## print datasets
tbl_FAT_order
tbl_INJ_order 

```

Sorted data by weather event and summed the number of property and crop damage. Selected values greater than zero and transformed property/crop damage values to account for exponent value in order to correctly sum damage data. Ordered the data from highest to lowest. 

```{r, cache=TRUE}

## Economic Damage -------------------------------------------------------------------------

## filter property damage OR crop damage values greater than 0
storm_data_tbl_econ <- filter(storm_data_tbl, PROPDMG > 0 | CROPDMG > 0)

## *** Replace exp abbreviation column with numerical values ***
## H-hundreds, M-millions, B-billions, K-thousands
## 0-9 was treated as power of 10, 10^X 
## -,+,? characters were ignored due to unknown values, changed to NA value
## blank value was treated as 10^0 = 1

## property damage exp column
storm_data_tbl_econ$PROPDMGEXP[ storm_data_tbl_econ$PROPDMGEXP  == "H" ] <- 100
storm_data_tbl_econ$PROPDMGEXP[ storm_data_tbl_econ$PROPDMGEXP  == "K" ] <- 1000 
storm_data_tbl_econ$PROPDMGEXP[ storm_data_tbl_econ$PROPDMGEXP  == "M" ] <- 1000000 
storm_data_tbl_econ$PROPDMGEXP[ storm_data_tbl_econ$PROPDMGEXP  == "B" ] <- 1000000000 
storm_data_tbl_econ$PROPDMGEXP[ storm_data_tbl_econ$PROPDMGEXP  == "2" ] <- 100 
storm_data_tbl_econ$PROPDMGEXP[ storm_data_tbl_econ$PROPDMGEXP  == "3" ] <- 1000
storm_data_tbl_econ$PROPDMGEXP[ storm_data_tbl_econ$PROPDMGEXP  == "4" ] <- 10000
storm_data_tbl_econ$PROPDMGEXP[ storm_data_tbl_econ$PROPDMGEXP  == "5" ] <- 100000
storm_data_tbl_econ$PROPDMGEXP[ storm_data_tbl_econ$PROPDMGEXP  == "6" ] <- 1000000 
storm_data_tbl_econ$PROPDMGEXP[ storm_data_tbl_econ$PROPDMGEXP  == "7" ] <- 10000000 
storm_data_tbl_econ$PROPDMGEXP[ storm_data_tbl_econ$PROPDMGEXP  == "8" ] <- 100000000 
storm_data_tbl_econ$PROPDMGEXP[ storm_data_tbl_econ$PROPDMGEXP  == "9" ] <- 1000000000 
storm_data_tbl_econ$PROPDMGEXP[ storm_data_tbl_econ$PROPDMGEXP  == "0" ] <- 1
storm_data_tbl_econ$PROPDMGEXP[ storm_data_tbl_econ$PROPDMGEXP  == "" ] <- 1
storm_data_tbl_econ$PROPDMGEXP[ storm_data_tbl_econ$PROPDMGEXP  == "-" ] <- NA
storm_data_tbl_econ$PROPDMGEXP[ storm_data_tbl_econ$PROPDMGEXP  == "+" ] <- NA

## crop damage exp column
storm_data_tbl_econ$CROPDMGEXP[ storm_data_tbl_econ$CROPDMGEXP  == "H" ] <- 100
storm_data_tbl_econ$CROPDMGEXP[ storm_data_tbl_econ$CROPDMGEXP  == "K" ] <- 1000 
storm_data_tbl_econ$CROPDMGEXP[ storm_data_tbl_econ$CROPDMGEXP  == "M" ] <- 1000000 
storm_data_tbl_econ$CROPDMGEXP[ storm_data_tbl_econ$CROPDMGEXP  == "B" ] <- 1000000000 
storm_data_tbl_econ$CROPDMGEXP[ storm_data_tbl_econ$CROPDMGEXP  == "0" ] <- 1
storm_data_tbl_econ$CROPDMGEXP[ storm_data_tbl_econ$CROPDMGEXP  == "" ] <- 1
storm_data_tbl_econ$CROPDMGEXP[ storm_data_tbl_econ$CROPDMGEXP  == "?" ] <- NA

## group by weather type, EVTYPE 
by_EVTYPE_econ <- group_by(storm_data_tbl_econ, EVTYPE)

## sum property and crop damage, seperate and combined
tbl_econ_PROPDMG <- summarize(by_EVTYPE_econ, 
                     "TOTAL.PROPDMG" = sum(PROPDMG*as.numeric(PROPDMGEXP)))
tbl_econ_CROPDMG <- summarize(by_EVTYPE_econ, 
                     "TOTAL.CROPDMG" = sum(CROPDMG*as.numeric(CROPDMGEXP)))
tbl_econ_TOTAL <- summarize(by_EVTYPE_econ, 
                     "TOTAL.DAMAGE" = sum(PROPDMG*as.numeric(PROPDMGEXP) + CROPDMG*as.numeric(CROPDMGEXP)) ) 

## arrange data in decending order
tbl_econ_PROPDMG_order <- arrange(tbl_econ_PROPDMG, 
                         desc(TOTAL.PROPDMG))
tbl_econ_CROPDMG_order <- arrange(tbl_econ_CROPDMG, 
                         desc(TOTAL.CROPDMG))
tbl_econ_TOTAL_order <- arrange(tbl_econ_TOTAL, 
                         desc(TOTAL.DAMAGE))

## print datasets
tbl_econ_PROPDMG_order
tbl_econ_CROPDMG_order
tbl_econ_TOTAL_order

```

## Results 

***Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?***

```{r figure1}

## make plot weather event vs. fatalities and injuries, 
library(ggplot2)

## first panel - fatalities 
plot1 <- ggplot(tbl_FAT_order[1:10,], aes(y=(TOTAL.FATALITIES/10^3), x=reorder(EVTYPE, TOTAL.FATALITIES))) + geom_bar(stat="identity") + coord_flip() + theme_bw() + labs(title = "TOP 10 WEATHER EVENTS: FATALITIES", x = "", y = "FATALITIES (Thousands, 10^3)")
plot1

## second panel - injuries
plot2 <- ggplot(tbl_INJ_order[1:10,], aes(y=(TOTAL.INJURIES/10^3), x=reorder(EVTYPE, TOTAL.INJURIES))) + geom_bar(stat="identity") + coord_flip() + theme_bw() + labs(title = "TOP 10 WEATHER EVENTS: INJURIES", x = "", y = "INJURIES (Thousands, 10^3)")
plot2
```

FIGURE 1: TOP 10 WEATHER EVENTS: FATALITIES AND INJURIES. (A) The top 5 fatality-causing events were TORNADO (5633), EXCESSIVE HEAT (1903), FLASH FLOOD (978), HEAT (937), and LIGHTNING (816). (B) The top 5 injury-causing events were TORNADO (91346), TSTM WIND (6957), FLOOD (6789), EXCESSIVE HEAT (6525), and LIGHTNING (5230). Tornados were the top of both categories by a significant margin.  
  
CONCLUSION: For greatest population health impact, future work/research/preparedness/aid should focus on Tornados. 
  
***Across the United States, which types of events have the greatest economic consequences?***  

```{r figure2}

## make plot weather event vs. damage, 3 panel plot ------------

## first panel - property damange 
plot3 <- ggplot(tbl_econ_PROPDMG_order[1:10,], aes(y=(TOTAL.PROPDMG/10^9), x=reorder(EVTYPE, TOTAL.PROPDMG))) + geom_bar(stat="identity") + coord_flip() + theme_bw() + labs(title = "TOP 10 WEATHER EVENTS: ECONOMIC COSTS", x = "", y = "PROPERTY DAMAGE ($ Billons, 10^9)")
plot3

## second panel - crop damage
plot4 <- ggplot(tbl_econ_CROPDMG_order[1:10,], aes(y=(TOTAL.CROPDMG/10^9), x=reorder(EVTYPE, TOTAL.CROPDMG))) + geom_bar(stat="identity") + coord_flip() + theme_bw() + labs(title = "TOP 10 WEATHER EVENTS: ECONOMIC COSTS", x = "", y = "CROP DAMAGE ($ Billons, 10^9)")
plot4

## third panel - total damage (PROPERTY + CROP)
plot5 <- ggplot(tbl_econ_TOTAL_order[1:10,], aes(y=(TOTAL.DAMAGE/10^9), x=reorder(EVTYPE, TOTAL.DAMAGE))) + geom_bar(stat="identity") + coord_flip() + theme_bw() + labs(title = "TOP 10 WEATHER EVENTS: ECONOMIC COSTS", x = "", y = "TOTAL DAMAGE (PROPERTY + CROP) ($ Billons, 10^9)")
plot5

```

FIGURE 2: TOP 10 WEATHER EVENTS: ECONOMIC COSTS. (A) The top 5 property damage causing events were FLOOD ($1.447e+11), HURRICANE/TYPHOON ($6.931e+10), STORM SURGE ($4.332e+10), FLASH FLOOD ($1.682e+10), and HAIL ($1.574e+10). (B) The top 5 crop damage causing events were DROUGHT ($1.397e+10), FLOOD ($5.662e+09), RIVER FLOOD ($5.029e+09), ICE STORM ($5.022e+09), and HAIL ($3.026e+09). (C) Total damage (property + crop). Floods and droughts were the top events by significant margins; however, the magnitude of property damage was significantly greater than crop damage. 

CONCLUSION: For greatest economic impact, future work/research/preparedness/aid should focus on Floods and Droughts. 
